Life in the canopy: Community trait assessments reveal substantial functional diversity among fern epiphytes

### 2020-02-19 ms/SI.rmd
Use a function to write out SI tables

# Load analysis results
loadd(list = c(
  "taxonomic_data_for_si",
  "trait_data_for_si",
  "climate_ancova_results",
  "phylosig_binary_traits_strict",
  "correlated_evo_test_strict",
  "best_fit_div_models_moran",
  "func_comm_struc_combined"), 
  cache = canopy_cache)

@@

# Set up table headers
si_table_header <- c(
  "New Phytologist Supporting Information",
  "Authors: Joel H. Nitta, James E. Watkins, Jr., and Charles C. Davis" %>% wrap_quotes,
  "Article acceptance date: 5 February 2020"
)
```

@@

'taxon code' is the name used for that taxon in analyses. 'informal variety' is a number assigned to infraspecific taxa that are morphologically and genetically distinct, but lack formal names distinguishing them. 'name source' indicates the source of the scientific name as follows: IPNI = International Plant Names Index (https://www.ipni.org/), TROPICOS = Missouri Botanical Garden taxonomic database (https://www.tropicos.org/), NCBI = National Center for Biotechnology Information taxonomic database (https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi), Murdock and Smith 2003 = Murdock & Smith (2003). 'NA' indicates not applicable.

```{r species-list}
# Set header and footer
header <- c(
  si_table_header,
  s_table_nums("species-list") %>% wrap_quotes
)

# Wrap footer at ca. 100 characters per line. Each line needs to be an element of the
# character vector.
footer <- "'taxon code' is the name used for that taxon in analyses. 'informal variety' is a number assigned to infraspecific taxa that are morphologically and genetically distinct, but lack formal names distinguishing them. 'name source' indicates the source of the scientific name as follows: IPNI = International Plant Names Index (https://www.ipni.org/), TROPICOS = Missouri Botanical Garden taxonomic database (https://www.tropicos.org/), NCBI = National Center for Biotechnology Information taxonomic database (https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi), Murdock and Smith 2003 = Murdock & Smith (2003). 'NA' indicates not applicable." %>% 
  str_wrap(width = 100) %>%
  str_split("\\n") %>%
  unlist %>%
  purrr::map_chr(wrap_quotes)

# Write out table with metadata
write_csv_with_meta(
  taxonomic_data_for_si, 
  here::here("results/Table_S1.csv"), 
  header,
  footer)
```

@@

SLA, specific leaf area. Continuous measurements shown as mean, standard deviation (sample size). 'taxon code' is the name used for that taxon in analyses (see Table S1). For data source, L = lab observation of gametophyte cultured from spore, F = field observation of gametophyte subsequently identified using DNA barcoding, T = morphology inferred based on taxonomy, M = measurement of a field-collected specimen; cited references provided at end of SI. 'NA' indicates missing data.

```{r trait_data}
# Set header and footer
header <- c(
  si_table_header,
  s_table_nums("trait-values") %>% wrap_quotes
)

footer <- "SLA, specific leaf area. Continuous measurements shown as mean, standard deviation (sample size). 'taxon code' is the name used for that taxon in analyses (see Table S1). For data source, L = lab observation of gametophyte cultured from spore, F = field observation of gametophyte subsequently identified using DNA barcoding, T = morphology inferred based on taxonomy, M = measurement of a field-collected specimen; cited references provided at end of SI. 'NA' indicates missing data." %>% 
  str_wrap(width = 100) %>%
  str_split("\\n") %>%
  unlist %>%
  purrr::map_chr(wrap_quotes)

# Write out table with metadata
write_csv_with_meta(
  trait_data_for_si, 
  here::here("results/Table_S2.csv"), 
  header,
  footer)
```

@@

d.f., degrees of freedom; MS, mean squares. Response variables include mean, minimum ('min'), maximum ('max'), and standard deviation ('sd') of temperature ('temp', °C) and vapor pressure deficit ('vpd', kPa). 'NA' indicates not applicable.

```{r climate-ancova}
# Format table
climate_ancova_results_formatted <-
climate_ancova_results %>% 
  mutate(
    p.value = signif(p.value, 2) %>% format.pval(., digits = 1),
    meansq = round_t(meansq, 2),
    term = case_when(
      term == "habit" ~ "Growth habit",
      term == "el" ~ "Elevation",
      term == "habit:el" ~ "Growth habit × elevation",
      TRUE ~ term
    )
  ) %>%
  select(`Response variable` = variable, `Source of variation` = term, `d.f.` = df, MS = meansq, P = p.value)

# Set header and footer
header <- c(
  si_table_header,
  s_table_nums("ancova") %>% wrap_quotes
)

footer <- "d.f., degrees of freedom; MS, mean squares. Response variables include mean, minimum ('min'), maximum ('max'), and standard deviation ('sd') of temperature ('temp', °C) and vapor pressure deficit ('vpd', kPa). 'NA' indicates not applicable." %>% wrap_quotes

# Write out table with metadata
write_csv_with_meta(
  climate_ancova_results_formatted, 
  here::here("results/Table_S3.csv"), 
  header,
  footer)
```

@@

@@

```{r phylosig-binary-strict}
# Format table
phylosig_binary_traits_strict_formatted <-
  phylosig_binary_traits_strict %>%
  mutate(D = round_t(D, 2)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("habit", "gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Epiphytic growth", "Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait) %>%
  select(Trait = trait, `Number of presences` = num_present, `Number of absences` = num_absent, D,
         P_rnd = prob_random, P_bm = prob_brownian)

# Set header and footer
header <- c(
  si_table_header,
  s_table_nums("phylosig-binary-strict") %>% wrap_quotes
)

footer <- "P_rnd, probability of obtaining the observed value of D under random trait evolution; P_bm, probability of obtaining the observed value of D under simulation of trait evolution by Brownian motion." %>%
  str_wrap(width = 100) %>%
  str_split("\\n") %>%
  unlist %>%
  purrr::map_chr(wrap_quotes)

# Write out table with metadata
write_csv_with_meta(
  phylosig_binary_traits_strict_formatted, 
  here::here("results/Table_S4.csv"), 
  header,
  footer)
```

@@

LL, log likelihood. Each gametophyte trait was coded as a binary trait and tested for correlated evolution with growth habit (epiphytic vs. terrestrial). Lower log likelihood for the dependent model indicates correlation with growth habit.

```{r corr-evo-strict}
# Format table
correlated_evo_test_strict_formatted <-
  correlated_evo_test_strict %>%
  mutate(trait = factor(
    trait, 
    levels = c("gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Gemmae", "Glands", "Hairs", "Morphotype"))) %>%
  arrange(trait) %>%
  select(
    Trait = trait,
    `LL (independent model)` = logL_indep,
    `LL (dependent model)` = logL_dep,
    `Likelihood ratio` = likelihood_ratio,
    `P` = pval
  )

# Set header and footer
header <- c(
  si_table_header,
  s_table_nums("corr-evo-strict") %>% wrap_quotes
)

footer <- "LL, log likelihood. Each gametophyte trait was coded as a binary trait and tested for correlated evolution with growth habit (epiphytic vs. terrestrial). Lower log likelihood for the dependent model indicates correlation with growth habit." %>%
  str_wrap(width = 100) %>%
  str_split("\\n") %>%
  unlist %>%
  purrr::map_chr(wrap_quotes)

# Write out table with metadata
write_csv_with_meta(
  correlated_evo_test_strict_formatted, 
  here::here("results/Table_S5.csv"), 
  header,
  footer)
```

**`r s_table("model-fit")`** `r s_table_cap("model-fit")` (see separate file)

MPDphy, standard effect size of mean phylogenetic distance; MNTDphy, standard effect size of mean nearest taxon distance; MPDfunc, standard effect size of mean functional distance; MNTDfunc, standard effect size of mean nearest functional distance; VPD, vapor pressure deficit, SD, standard deviation; AICc, corrected Akaike Information Criterion; P_moran, \pval\ value for the null hypothesis that there is spatial autocorrelation in the residuals as measured with Moran's *I*. Models within delta AICc 3 of the best-fitting model shown. Trait names refer to community-weighted mean values (see Methods). Frond length and width were highly correlated with rhizome diameter and not shown. For details on calculations of MPD and MNTD, see Methods.

```{r model-fit}
# Format table
best_fit_div_models_moran_formatted <-
best_fit_div_models_moran %>%
  # Drop frond length and width (highly correlated with rhizome dia.)
  filter(resp_var %in% c("ntaxa", "mpd.obs.z.phy", "mntd.obs.z.phy", 
                   "mpd.obs.z.func", "mntd.obs.z.func",
                   "dissection", "sla", "stipe", "rhizome", "pinna")) %>%
  mutate(resp_var = factor(
        resp_var, 
        levels = c("ntaxa", "mpd.obs.z.phy", "mntd.obs.z.phy", 
                   "mpd.obs.z.func", "mntd.obs.z.func",
                   "dissection", "sla", "stipe", "rhizome", "pinna"))) %>%
  mutate(
    resp_var = lvls_revalue(
        resp_var, 
        new_levels = c("Richness", "MPDphy", "MNTDphy", 
                       "MPDfunc", "MNTDfunc",
                       "Dissection", "SLA", "Stipe length", "Rhizome diam.", "Pinna no.")) 
  ) %>%
  arrange(resp_var, delta_AICc) %>%
  mutate(
    modname = modname %>%
      str_replace_all("\\+", " \\+ ") %>%
      str_replace_all("\\.by\\.", " x ") %>%
      str_replace_all("habit", "Growth habit") %>%
      str_replace_all("temp_mean", "Mean temp.") %>%
      str_replace_all("temp_sd", "SD temp.") %>%
      str_replace_all("temp_max", "Max. temp.") %>%
      str_replace_all("vpd_mean", "Mean VPD") %>%
      str_replace_all("vpd_sd", "SD VPD") %>%
      str_replace_all("vpd_min", "Min. VPD")
  ) %>% 
  select(
    `Metric` = resp_var,
    `Model` = modname,
    AICc,
    `R2` = r2,
    delta_AICc,
    `Moran's I` = morans_I,
    `P_moran` = morans_I_pval
  )

# Set header and footer
header <- c(
  si_table_header,
  s_table_nums("model-fit") %>% wrap_quotes
)

footer <- "MPDphy, standard effect size of mean phylogenetic distance; MNTDphy, standard effect size of mean nearest taxon distance; MPDfunc, standard effect size of mean functional distance; MNTDfunc, standard effect size of mean nearest functional distance; VPD, vapor pressure deficit, SD, standard deviation; AICc, corrected Akaike Information Criterion; P_moran, P value for the null hypothesis that there is spatial autocorrelation in the residuals as measured with Moran's I. Models within delta AICc 3 of the best-fitting model shown. Trait names refer to community-weighted mean values (see Methods). Frond length and width were highly correlated with rhizome diameter and not shown. For details on calculations of MPD and MNTD, see Methods." %>%
  str_wrap(width = 100) %>%
  str_split("\\n") %>%
  unlist %>%
  purrr::map_chr(wrap_quotes)

# Write out table with metadata
write_csv_with_meta(
  best_fit_div_models_moran_formatted, 
  here::here("results/Table_S6.csv"), 
  header,
  footer)

@@

>>>>

# Load analysis results
loadd(list = c(
  "taxonomic_data_for_si",
  "trait_data_for_si",
  "climate_ancova_results",
  "phylosig_binary_traits_strict",
  "correlated_evo_test_strict",
  "best_fit_div_models_moran",
  "func_comm_struc_combined",
  "climate_model_summaries",
  "climate_models",
  "div_el_models",
  "div_el_model_summaries",
  "div_el_model_parameter_summaries",
  "cwsd_scatterplot"), 
  cache = canopy_cache)

@@

# Load functions
source(here::here("code/functions.R"))

# Set up main table header
si_table_header <- c(
  "New Phytologist Supporting Information",
  "Authors: Joel H. Nitta, James E. Watkins, Jr., and Charles C. Davis" %>% wrap_quotes,
  "Article acceptance date: 5 February 2020"
)

@@

```{r species-list}
species_list_footer <- "'Taxon code' is the name used for that taxon in analyses. 'Informal variety' is a number assigned to infraspecific taxa that are morphologically and genetically distinct, but lack formal names distinguishing them. 'Name source' indicates the source of the scientific name as follows: IPNI = International Plant Names Index (https://www.ipni.org/), TROPICOS = Missouri Botanical Garden taxonomic database (https://www.tropicos.org/), NCBI = National Center for Biotechnology Information taxonomic database (https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi), Murdock and Smith 2003 = Murdock & Smith (2003). The genus in 'Taxon code' may differ from the 'Genus' column for names whose preferred synonymy has changed since data were originally recorded. 'NA' indicates not applicable."

write_si_table(
  table_id = "species-list", 
  main_header = si_table_header,
  table = taxonomic_data_for_si,
  table_footer = species_list_footer
)
```

`r species_list_footer`

@@

```{r trait_data}
trait_values_footer <- "SLA, specific leaf area. Continuous measurements shown as mean, standard deviation (sample size). 'taxon code' is the name used for that taxon in analyses (see Table S1). For data source, L = lab observation of gametophyte cultured from spore, F = field observation of gametophyte subsequently identified using DNA barcoding, T = morphology inferred based on taxonomy, M = measurement of a field-collected specimen; cited references provided at end of SI. 'NA' indicates missing data."

write_si_table(
  table_id = "trait-values", 
  table = trait_data_for_si,
  main_header = si_table_header,
  table_footer = trait_values_footer
)
```

`r trait_values_footer`

@@

```{r ancova}
# Format table
ancova_table <-
climate_ancova_results %>% 
  mutate(
    p.value = signif(p.value, 2) %>% format.pval(., digits = 1),
    meansq = round_t(meansq, 2),
    term = case_when(
      term == "habit" ~ "Growth habit",
      term == "el" ~ "Elevation",
      term == "habit:el" ~ "Growth habit × elevation",
      TRUE ~ term
    )
  ) %>%
  select(`Response variable` = variable, `Source of variation` = term, `d.f.` = df, MS = meansq, P = p.value)

ancova_footer <- "d.f., degrees of freedom; MS, mean squares. Response variables include mean, minimum ('min'), maximum ('max'), and standard deviation ('sd') of temperature ('temp', °C) and vapor pressure deficit ('vpd', kPa). 'NA' indicates not applicable."

write_si_table(
  table_id = "ancova", 
  table = ancova_table,
  main_header = si_table_header,
  table_footer = ancova_footer
)
```

`r ancova_footer`

**`r s_table("linear-fit")`** `r s_table_cap("linear-fit")` (see separate file)

```{r linear-fit}
# Combine summaries of models for diversity metrics
div_summaries <-
  div_el_model_summaries %>%
  left_join(
    select(div_el_models,
           var, model_type, delta_AICc, morans_I, I_pval), 
    by = c("var", "model_type")
  )

# Combine summaries of models for climate metrics
climate_summaries <-
  climate_model_summaries %>%
  left_join(
    select(climate_models,
           var, model_type, delta_AICc, morans_I, I_pval), 
    by = c("var", "model_type")
  )

# Combine summaries of diversity and climate metrics, reformat model names
combined_summaries <-
  bind_rows(div_summaries, climate_summaries) %>%
  mutate(
    model_type = model_type %>% 
      str_replace_all("el_only", "Elevation") %>%
      str_replace_all("both", "Elevation + Growth habit") %>%
      str_replace_all("habit_only", "Growth habit") %>%
      str_replace_all("interaction", "Elevation + Growth habit + Elevation x Growth habit")
    )

# Reformat variable names and put in preferred order by joining
linear_fit_table <-
tribble(
  ~var, ~Response,
  "temp_max", "Max. temp.",
  "temp_mean", "Mean temp.",
  "vpd_mean", "Mean VPD",
  "vpd_min", "Min. VPD",
  "temp_sd", "SD temp.",
  "vpd_sd", "SD VPD",
  "ntaxa", "Richness",
  "mpd.obs.z.phy", "MPDphy",
  "mntd.obs.z.phy", "MNTDphy",
  "mpd.obs.z.func", "MPDfunc",
  "mntd.obs.z.func", "MNTDfunc",
  "dissection", "Dissection",
  "length", "Frond length",
  "pinna", "Pinna no.",
  "rhizome", "Rhizome diam.",
  "stipe", "Stipe length",
  "sla", "SLA",
  "width", "Frond width"
) %>%
  left_join(combined_summaries, by = "var") %>%
  select(
    Response,
    Model = model_type,
    AICc,
    f = statistic,
    `Adjusted R2` = adj.r.squared,
    P = p.value,
    `d.f.` = df,
    delta_AICc,
    `Moran's I` = morans_I,
    P_moran = I_pval
  )

linear_fit_footer <- "MPDphy, standard effect size of mean phylogenetic distance; MNTDphy, standard effect size of mean nearest taxon distance; MPDfunc, standard effect size of mean functional distance; MNTDfunc, standard effect size of mean nearest functional distance; VPD, vapor pressure deficit, SD, standard deviation; SLA, specific leaf area; AICc, corrected Akaike Information Criterion; d.f., degrees of freedom; P_moran, P value for the null hypothesis that there is spatial autocorrelation in the residuals as measured with Moran's I. Trait names refer to community-weighted mean values (see Methods). For details on calculations of MPD and MNTD, see Methods. 'NA' indicates not applicable."

write_si_table(
  table_id = "linear-fit", 
  table = linear_fit_table,
  main_header = si_table_header,
  table_footer = linear_fit_footer
)
```

`r linear_fit_footer`

@@

```{r phylosig-binary-strict}
# Format table
phylosig_binary_strict_table <-
  phylosig_binary_traits_strict %>%
  mutate(D = round_t(D, 2)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("habit", "gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Epiphytic growth", "Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait) %>%
  select(Trait = trait, `Number of presences` = num_present, `Number of absences` = num_absent, D,
         P_rnd = prob_random, P_bm = prob_brownian)

phylosig_binary_strict_footer <- "P_rnd, probability of obtaining the observed value of D under random trait evolution; P_bm, probability of obtaining the observed value of D under simulation of trait evolution by Brownian motion."

write_si_table(
  table_id = "phylosig-binary-strict", 
  table = phylosig_binary_strict_table,
  main_header = si_table_header,
  table_footer = phylosig_binary_strict_footer
)
```

@@

@@

```{r corr-evo-strict}
# Format table
corr_evo_strict_table <-
  correlated_evo_test_strict %>%
  mutate(trait = factor(
    trait, 
    levels = c("gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Gemmae", "Glands", "Hairs", "Morphotype"))) %>%
  arrange(trait) %>%
  select(
    Trait = trait,
    `LL (independent model)` = logL_indep,
    `LL (dependent model)` = logL_dep,
    `Likelihood ratio` = likelihood_ratio,
    `P` = pval
  )

corr_evo_strict_footer <- "LL, log likelihood. Each gametophyte trait was coded as a binary trait and tested for correlated evolution with growth habit (epiphytic vs. terrestrial). Lower log likelihood for the dependent model indicates correlation with growth habit."

write_si_table(
  table_id = "corr-evo-strict", 
  table = corr_evo_strict_table,
  main_header = si_table_header,
  table_footer = corr_evo_strict_footer
)
```

`r corr_evo_strict_footer`

**`r s_table("gam-fit")`** `r s_table_cap("gam-fit")` (see separate file)

```{r gam-fit}
# Format table
gam_fit_table <-
best_fit_div_models_moran %>%
  # Drop frond length and width (highly correlated with rhizome dia.)
  filter(resp_var %in% c("ntaxa", "mpd.obs.z.phy", "mntd.obs.z.phy", 
                   "mpd.obs.z.func", "mntd.obs.z.func",
                   "dissection", "sla", "stipe", "rhizome", "pinna")) %>%
  mutate(resp_var = factor(
        resp_var, 
        levels = c("ntaxa", "mpd.obs.z.phy", "mntd.obs.z.phy", 
                   "mpd.obs.z.func", "mntd.obs.z.func",
                   "dissection", "sla", "stipe", "rhizome", "pinna"))) %>%
  mutate(
    resp_var = lvls_revalue(
        resp_var, 
        new_levels = c("Richness", "MPDphy", "MNTDphy", 
                       "MPDfunc", "MNTDfunc",
                       "Dissection", "SLA", "Stipe length", "Rhizome diam.", "Pinna no.")) 
  ) %>%
  arrange(resp_var, delta_AICc) %>%
  mutate(
    modname = modname %>%
      str_replace_all("\\+", " \\+ ") %>%
      str_replace_all("\\.by\\.", " x ") %>%
      str_replace_all("habit", "Growth habit") %>%
      str_replace_all("temp_mean", "Mean temp.") %>%
      str_replace_all("temp_sd", "SD temp.") %>%
      str_replace_all("temp_max", "Max. temp.") %>%
      str_replace_all("vpd_mean", "Mean VPD") %>%
      str_replace_all("vpd_sd", "SD VPD") %>%
      str_replace_all("vpd_min", "Min. VPD")
  ) %>% 
  select(
    `Metric` = resp_var,
    `Model` = modname,
    AICc,
    `R2` = r2,
    delta_AICc,
    `Moran's I` = morans_I,
    `P_moran` = morans_I_pval
  )

gam_fit_footer <- "MPDphy, standard effect size of mean phylogenetic distance; MNTDphy, standard effect size of mean nearest taxon distance; MPDfunc, standard effect size of mean functional distance; MNTDfunc, standard effect size of mean nearest functional distance; VPD, vapor pressure deficit, SD, standard deviation; AICc, corrected Akaike Information Criterion; P_moran, P value for the null hypothesis that there is spatial autocorrelation in the residuals as measured with Moran's I. Models within delta AICc 3 of the best-fitting model shown. Trait names refer to community-weighted mean values (see Methods). Frond length and width were highly correlated with rhizome diameter and not shown. For details on calculations of MPD and MNTD, see Methods."

write_si_table(
  table_id = "gam-fit", 
  table = gam_fit_table,
  main_header = si_table_header,
  table_footer = gam_fit_footer
)

@@

`r gam_fit_footer`

**`r s_table("spatial-sla")`** `r s_table_cap("spatial-sla")` (see separate file)

```{r spatial-sla}
# Format table
spatial_sla_table <-
div_el_model_parameter_summaries %>%
  filter(var == "sla") %>%
  select(term, Coefficient = estimate, SE = std.error, t = t_value) %>%
  mutate(term = term %>%
           str_replace_all("habitterrestrial", "Growth habit") %>%
           str_replace_all("el", "Elevation")) %>%
  rename(Term = term)

spatial_sla_footer <- "A Mátern spatial correlation structure was included in the model to account for spatial autocorrelation (see Methods)."

write_si_table(
  table_id = "spatial-sla", 
  table = spatial_sla_table,
  main_header = si_table_header,
  table_footer = spatial_sla_footer
)

`r spatial_sla_footer`
